Historical Soil Organic Carbon Budget

### 2020-06-26 SOCbudgetPaper/methods.Rmd
update method part + smaller stuff in other section and result figure

<!-- Soil carbon dynamics are modeled using a yearly based SOC model and carbon input and management data. -->

@@

Following the tier 2 approach of the refinement of IPCC guidelines vol. 4 (@ipcc_2019_2019), we estimate global land-use type specific soil organic carbon (SOC) stocks for cropland and natural vegetation on half-degree resolution from 1965 to 2010. We assume the actual SOC state converges towards a stable steady state, that itself is changing over time and space depending on biophysical, climatic and agronomic conditions. Therefore we conduct the following three steps within each yearly timestep: 
(1) We calculate annual land-use (sub-)type specific steady states and decay rates for SOC stocks, 
(2) We account for land conversion by transferring SOC between land-use types and 
(3) We estimate SOC stocks based on the stocks of the previous time period, the steady state stocks and the decay rate.

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{eq}$ are given by 
\begin{equation}
SOC^{eq}_t =\frac{C^{\textrm{in}}_t}{k_t}
\label{eq:inoutflow}
\end{equation}
with $C^{\textrm{in}}$ being carbon inputs to the soil and $k$ denoting the soil organic carbon decay rate. We use for our calculations the steady-state method of the refinement of the IPCC guidelines vol. 4 (@ipcc_2019_2019) for mineral soils, which assume three soil carbon sub-pools (active, slow and passive) and entangled dynamics between them. Annual carbon inflow to each sub-pool (see \@ref(sec:carboninputs)) and annual decay rates (see \@ref(sec:tier2)) of each sub-pool are still the key components to determining steady-state SOC stocks.

<!-- BB: I would include the t, also to show that the steady-state is time-dependent. -->

@@

We account for different carbon input sources depending on the land-use type (see table \@ref(tab:datasourceinputs)).
Name here the inputs, and refer to the respective sections, dont just refer to the table.
Following the IPCC methodology carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content (see \@ref(ipcc_2019_2019)). For each component the sum over all carbon input sources is allocated to the respective SOC sub-pools via transfer coefficients. This implies that not only the amount of carbon, but also their structural composition is determining the effective inflow. Data sources for all considered carbon inputs as well as for lignin and nitrogen content can be found in table \@ref(tab:datasourceinputs).

@@

@@

\begin{equation}
SOC_{t} = SOC_{t-1} + (SOC^{eq}_{t} - SOC_{t-1}) \cdot k_{t} \cdot 1\unit{a} \qquad \forall lut, sub, i.
\label{eq:steadystate}
\end{equation}

@@

\begin{equation}
SOC_{t} = SOC_{t-1} - \underbrace{SOC_{t-1} \cdot k_{t} \cdot 1\unit{a}}_{\text{outflow}} + \overbrace{SOC^{eq}_{t} \cdot k_{t} \cdot 1\unit{a}}^{\text{input (using equation \ref(eq:inoutflow))}}  \qquad \forall lut, sub, i.
\label{eq:steadystate}
\end{equation}

@@

Additionally to the tier 2 approach of the refinement of IPCC guidelines vol. 4 (@ipcc_2019_2019), we also estimate SOC pools using the IPCC tier 1 approach of IPCC guidelines vol. 4 (@ipcc_2006_2006) for comparison. Here, stocks are estimated via stock change factors given by the IPCC for the topsoil (0-30 cm) and based on a review of measurement data. The factors differentiate  different crop and management systems reflecting different dynamics under changed in- and outflows without explicitly tracking these. The SOC stocks as thus calulated

\begin{equation}
SOC_{\text{target}} = \sum_{c,s,i} SOC_{\text{ref}_{c,s,i}} \cdot F_{\text{LU}_{c,s,i}} \cdot F_{\text{MG}_{c,s,i}} \cdot F_{\text{I}_{c,s,i}} \cdot A_{c,s,i}
\label{eq:tier1}
\end{equation}

<!- also include an equation here -->
<!- even if there are just "copied" out of te guidelines so to say? -->
<!- more details will follow - how deep to go? -->

@@

Agricultural management data is estimated using the R library mrcommons ([cite git link]), which compiles country-specific FAO production and cropland statistics ([@FAOSTAT]) to a comprehensive and constistent data suite. <!-- BB: Isnt the whole calculation based on R libraries? - KK: Sure but I thought makig the relevenat packages visible here makes sense? --> 
The data is prepared in 5 year time steps from 1965 to 2010, which also restricts our analysis to this time span.
For all the following data, if not declared differently, we interpolate values linearly between the time steps.

@@

Land-use patterns are based on the Land-Use Harmonization 2 (LUH2, [@LUH2]) data set, which we aggregate from quarter degree to half degree resolution. We disaggregate the five different cropland subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17 crop groups, applying the relative shares for each gridcell based on the country and year specific area harvested shares of FAOSTAT data ([@FAOSTAT]) (see \@ref(append:Table_luh2fao2mag) for more  details on the crop type mapping). Land-use transitions are calculated as net area differences of the land-use data on half-degree.

### Crop, Crop Residues and Pasture Production	(300)

@@

#### Crop Residue Production
Crop residue production and management is based on a revised methodology of ([@bodirsky2012]) and will be explained in key aspects again due to its central role for soil carbon modelling. Starting from crop production estimates of the harvested organs and their respective crop area, we estimate above-ground (ag) and below-ground (bg) residual biomass using yield-dependent <!-- do not know what was meant with this anymore (why "yield-dependent"?). I guess that has to be explained again than -.- --> harvest indices and shoot:root ratios. We assume that all bg residues are recycled to the soil, whereas ag residues can be burned or harvested for other purposes such as feeding animals ([@weindl]), fuel or for material use.
    <!-- BB: One new thing is here that we use a doublecropping factor to account for lower harvest indices of multicropping vs a very high single-cropped yield - KK: ASk Benni again about this -->

##### Burned Residues
A fixed share of the ag residues is assumed to be burned on field depending on the per-capita income of the country. Following [@smil1999]) we assume 25% burn share for low-income countries according to worldbank definitions ($<\,1000\,\tfrac{USD}{yr}$), 15% for high-income  ($>\,10000\,\tfrac{USD}{yr}$ and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop type 80–90% of the residue carbon burned on the fields are lost within the combustion process ([@ipcc_2006_2006]).

##### Residue Usage
We compile out of our 17 crop groups, three used residue groups (straw, high-lignin and low-lignin residues) with additional demand for other purposes and one residues with no double use (see \@ref(append:Table_kcr2kres)).

Residue feed demand for five different livestock groups is based on country- and residue-group-specific feed basekts (see [@weindl]) taking available ag residual biomass as well as livestock productivity into account.

We estimate a material-use share for the straw residues group of 5% and a fuel-share of 10% for all used residues groups in low income countries according to worldbank definitions ($<\,1000\,\tfrac{USD}{yr}$). For high-income  ($>\,10000\,\tfrac{USD}{yr}$ no withdrawl for material or fuel use is assumend, leaving middle-income countries with linearly interpolate shares depending on their per-capita income.

The remaining ag residues as well as all bg residues are assumend to be recycled to the soil. 
We cut high recycling shares per hectar at the 95%-percentile to corrected for outliers.

#### Pasture Production
Using livestock production statistics as well as feed mix assumptions as describted in ([@weindl]) we estimating country specific pasture production. Following the same approach as for crop production we disaggregate and calibrate half-degree pasture production pattern from grass yields from LPJmL and pasture area and rangeland patterns ( (see \@ref(#sec:landuse))) to derive half-degree pasture production patterns.
<!-- BB: Oh, does that mean we assume that only the pasture grows that is being eaten? Thats not too much, is it? Maybe ask Isabelle for a better solution (e.g. direct pasture yields from LPJmL). And then the "pasture production" is just the part being removed by grazing. Does the IPCC specifiy something in respect to pastures?  - KK: Clarify with Benni.But actually pastrue production does not matter since just the patterns are used to distribute livestock. Anyway are word on the neglected SOC cycling on pasture compared to natveg has to come somewhere. -->

@@

<!-- Litter is coming from LPJmL in carbon units - transformation with 0.44 is done twice reverting the effect of the transformation -->

### Livestock Distribution and Manure Excretion	(300)

#### Livestock Distribution
To disaggregate country level FAOSTAT livestock production values to half-degree pattern, we use the following rule based assumptions which were inspired by the approach of [@gilbert] and uses feed basket assumptions based on a revised methodology of @weindl.
We differentiate ruminant and monogastric systems, as well as extensive and an intensive systems.
For ruminants, we use the country-level shares of gras within the feed baskets to split up pasture-fed, extensive systems from the rather intensive crop-fed livestock.
For extensive dairy and ruminant meat production we estimate, that livestock is located in proximity to pastures and rangelands and distribute manure excretion proportional to pasture production of all half-degree gridcell of a country. 
Intensive dairy and ruminant meat production is assumend to be located proportinal to crop production to have short transport distances for feed stuff.  <!-- yeah, actually was that the full idea here? -->

<!-- BB: What does not become clear to me: Do the extensive cows only eat pasture, and the intensive one only crops? Does that in the end mean we just redistirbute grazed pasture to pastures and crop-based feed to croplands proportional to production? Or is there more to it? - KK: Now acctually not.-->

For poultry, egg and monogastric meat production we use the per-capita income of the country to divide into intensive and extensive production systems. For low-income countries according to worldbank definitions (<1000 USD/yr), we assume extensive production systems. We locate them according to built-up areas shares based on the idea that these animals are held in households, subsistence or small-holder farming systems with a high labour per animal ratio. Intensive production is distributed within a country using the crop production share, assuming that feed availability is the most driving factor for livestock location.

#### Manure Excretion, Storage and Recycling
Manure production and management is based on a revised methodology of ([@bodirsky2012]) and will be explained in key aspects again due to its central role for soil carbon modelling. Based on the gridded livestock distribution we calculate excretions by estimating the nitrogen balance of the livestock system on the basis of comprehensive livestock feed baskets ([@weindl]), assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed C:N ratios (given by [(@ipcc_2019_2019)]). 
Depending on the feed system we assume manure to be handled in four different ways:
All manure orginated from pasture feed intake is excreted directly to pastures and rangelands (pasture grazing), deducting manure collected as fuel. Manure fuel shares are estimated using IPCC default values (@ippc_2006_2006). Whereas for low-income countries according to worldbank definitions (<1000 USD/yr), we adopt a share of 25% of crop residues in feed intake directly consumend and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries ($>\,10000\,\tfrac{USD}{yr}$, leaving middle-income countries with linearly interpolate shares depending on their per-capita income.
For all other feed items we assume the manure to be stored in animal waste management systems associated to animal houses. 
To estimate the carbon actually recycled to the soil, we account for carbon losses during storage and recycling shares in different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not recycled. For manure stored in different animal waste management system we compiled carbon loss rates partly depending on the nitrogen loss rates as specified in [@bodirsky2012] (see \@ref(append:Table_clossAWMS))).

@@

Tillage data sets of [Vera, others] together with rules are used to drive tillage effect on decay rates.

>>>>

We compiled calculations as open-source R packages available under github.com/pik-piam/mrcommons (management related functions) and github.com/pik-piam/mrSOCbudget (soil dynamic related functions), which are both based on github.com/pik-piam/madrat, a framework improving reproducibility and transparency in data processing.
In the following chapter we outine the most important relationships and assumptions. See table \@ref(append:subsection2mrfunctions) for further information on corresponding code within the R packages.

@@

Following the tier 2 approach of the refinement of IPCC guidelines vol. 4 (@ipcc_2019_2019), we estimate soil organic carbon (SOC) stocks for cropland and all other land represented by natural vegetation on half-degree resolution from 1975 to 2010. We assume the actual SOC state converges towards a stable steady state, that itself is changing over time and space depending on biophysical, climatic and agronomic conditions. Therefore we conduct the following three steps within each yearly timestep: 
(1) We calculate annual land-use (sub-)type specific steady states and decay rates for SOC stocks, 
(2) We account for land conversion by transferring SOC between land-use types and 
(3) We estimate SOC stocks based on the stocks of the previous time period, the steady state stocks and the decay rate.

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{eq}$ are given by 
\begin{equation}
SOC^{eq} =\frac{C^{\textrm{in}}}{k} \qquad\forall\quad i,t
\label{eq:inoutflow}
\end{equation}
with $C^{\textrm{in}}$ being carbon inputs to the soil, $k$ denoting the soil organic carbon decay rate; as well as $i$ representing grid cells indecies and $t$ years. We use for our calculations the steady-state method of the refinement of the IPCC guidelines vol. 4 (@ipcc_2019_2019) for mineral soils, which assume three soil carbon sub-pools (active, slow and passive) and entangled dynamics between them. Annual carbon inflow to each sub-pool (see \@ref(sec:carboninputs)) and annual decay rates (see \@ref(sec:tier2)) of each sub-pool are still the key components to determining steady-state SOC stocks.

@@

We account for different carbon input sources depending on our two distinguished land-use types (see table \@ref(tab:datasourceinputs)); in particular for recycled crop residues, below ground biomass of crops (for both see \@ref(sec:residues)) and recycled manure (see \ref(sec:livst_manure)) on croplands; and litterfall (\ref(@LPJmL4_1)) on natural vegetation.

Following the IPCC methodology carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content (see \@ref(ipcc_2019_2019)). For each component the sum over all carbon input sources is allocated to the respective SOC sub-pools via transfer coefficients. This implies that not only the amount of carbon, but also their structural composition is determining the effective inflow. Data sources for all considered carbon inputs as well as for lignin and nitrogen content can be found in table \@ref(tab:datasourceinputs).

@@

<!-- add references here for the table-->

@@

\begin{equation}
SOC_{t} = SOC_{t-1} + (SOC^{eq}_{t} - SOC_{t-1}) \cdot k_{t} \cdot 1\unit{a} \qquad \forall\quad\quad lut, sub, i.
\label{eq:steadystate}
\end{equation}

@@

\begin{equation}
SOC_{t} = SOC_{t-1} - \underbrace{SOC_{t-1} \cdot k_{t} \cdot 1\unit{a}}_{\text{outflow}} + \overbrace{SOC^{eq}_{t} \cdot k_{t} \cdot 1\unit{a}}^{\text{input (using equation \ref(eq:inoutflow))}}  \qquad \forall\quad lut, sub, i.
\label{eq:steadystate}
\end{equation}

@@

@@

We compile country-specific FAO production and cropland statistics ([@FAOSTAT]) to a comprehensive and constistent data suite. The data is prepared in 5 year time steps from 1965 to 2010, which also restricts our analysis together with the accounting for a spin up phase to time span from 1975 to 2010. For all the following data, if not declared differently, we interpolate values linearly between the time steps.

@@

Land-use patterns are based on the Land-Use Harmonization 2 (LUH2, [@LUH2]) data set, which we aggregate from quarter degree to half degree resolution. We disaggregate the five different cropland subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17 crop groups, applying the relative shares for each gridcell based on the country and year specific area harvested shares of FAOSTAT data ([@FAOSTAT]) (see \@ref(append:Table_luh2fao2mag) for more  details on the crop group mapping). Land-use transitions are calculated as net area differences of the land-use data on half-degree.

### Crop and Crop Residues (300) {#sec:residues}

@@

#### Crop Residue Production
Crop residue production and management is based on a revised methodology of ([@bodirsky2012]) and will be explained in key aspects again due to its central role for soil carbon modelling. Starting from crop production ($CP$) estimates of the harvested organs and their respective harvested crop area ($CA$), we estimate above-ground ($AGR$) and below-ground ($BGR$) residual biomass using crop group ($cg$) specific harvest indices ($HI$) and root:shoot ratios ($RS$) as follows

\begin{equation}
\begin{aligned}
AGR & = CP \cdot HI_{\textrm{slope}} + CA \cdot HI_{\textrm{intercept}}\qquad & \textrm{and} \\
BGR & = (CP + AGR) \cdot RS \qquad                                            & \forall\quad cg, i, t.
\label{eq:resbiomass}
\end{aligned}
\end{equation}

Following the IPCC guidelines, we split the harvest index into a yield and an area dependend fraction ([@ipcc_2006_2006]). Note that deviating from [@bodirsky2012] we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased amounts on fallow land. 
We assume that all BGR are recycled to the soil, whereas AGR can be burned or harvested for other purposes such as feeding animals ([@weindl]), fuel or for material use.

##### Burned Residues
A fixed share of the AGR is assumed to be burned on field depending on the per-capita income of the country. Following [@smil1999]) we assume 25% burn share for low-income countries according to worldbank definitions ($<\,1000\,\tfrac{USD}{yr}$), 15% for high-income  ($>\,10000\,\tfrac{USD}{yr}$ and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group 80–90% of the residue carbon burned on the fields are lost within the combustion process ([@ipcc_2006_2006]).

##### Residue Usage
We compile out of our 17 crop groups, three used residue groups (straw, high-lignin and low-lignin residues) with additional demand for other purposes and one residues with no double use (see \@ref(append:Table_kcr2kres)). Residue feed demand for five different livestock groups is based on country- and residue-group-specific feed basekts (see [@weindl]) taking available AGR biomass as well as livestock productivity into account. We estimate a material-use share for the straw residues group of 5% and a fuel-share of 10% for all used residues groups in low income countries according to worldbank definitions ($<\,1000\,\tfrac{USD}{yr}$). For high-income  ($>\,10000\,\tfrac{USD}{yr}$ no withdrawl for material or fuel use is assumend, leaving middle-income countries with linearly interpolate shares depending on their per-capita income. The remaining AGR as well as all BGR are assumend to be recycled to the soil. We cut high recycling shares at $10\tfrac{\unit{tC}}{\unit{ha}}$ to corrected for outliers.

@@

### Livestock Distribution and Manure Excretion	(300) {#sec:livst_manure}

We assume that manure is applied at its excretion place, leaving the livestock distribution the driving factor for the spatial pattern of manuring.

#### Livestock Distribution
To disaggregate country level FAOSTAT livestock production values to half-degree pattern, we use the following rule based assumptions which were inspired by the approach of [@gilbert] and uses feed basket assumptions based on a revised methodology of @weindl.We differentiate ruminant and monogastric systems, as well as extensive and an intensive systems.
For ruminants, we assume that livestock is located at the production spot sof the feed intake. We distingush between grazed pasture, which is converted into livestock products from extensive systems; and all other (crop-based) feed stuff, which we consider to be consumend in intensive systems.
For poultry, egg and monogastric meat production we use the per-capita income of the country to divide into intensive and extensive production systems. For low-income countries according to worldbank definitions (<1000 USD/yr), we assume extensive production systems. We locate them according to built-up areas shares based on the idea that these animals are held in households, subsistence or small-holder farming systems with a high labour per animal ratio. Intensive production is distributed within a country using again crop-based feed production share, assuming that feed availability is the most driving factor for livestock location.

#### Manure Excretion, Storage and Recycling
Manure production and management is based on a revised methodology of ([@bodirsky2012]) and will be explained in key aspects again due to its central role for soil carbon modelling. Based on the gridded livestock distribution we calculate excretions by estimating the nitrogen balance of the livestock system on the basis of comprehensive livestock feed baskets ([@weindl]), assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed C:N ratios (given by [(@ipcc_2019_2019)]). 
Depending on the feed system we assume manure to be handled in four different ways:
All manure orginated from pasture feed intake is excreted directly to pastures and rangelands (pasture grazing), deducting manure collected as fuel. Manure fuel shares are estimated using IPCC default values (@ippc_2006_2006). 
Whereas for low-income countries according to worldbank definitions (<1000 USD/yr), we adopt a share of 25% of crop residues in feed intake directly consumend and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries ($>\,10000\,\tfrac{USD}{yr}$, leaving middle-income countries with linearly interpolate shares depending on their per-capita income.
For all other feed items we assume the manure to be stored in animal waste management systems associated to animal houses. 
To estimate the carbon actually recycled to the soil, we account for carbon losses during storage and recycling shares in different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not recycled. For manure stored in different animal waste management system we compiled carbon loss rates partly depending on the nitrogen loss rates as specified in [@bodirsky2012] (see \@ref(append:Table_clossAWMS))). We cut high application shares at $10\tfrac{\unit{tC}}{\unit{ha}}$ to corrected for outliers.

@@

Additionally to the tier 2 approach of the refinement of IPCC guidelines vol. 4 (@ipcc_2019_2019) and the detailed analysis of management data coming with it, SOC changes can be estimated using the IPCC tier 1 approach of IPCC guidelines vol. 4 (@ipcc_2006_2006, @ipcc_2019_2019). Here, stocks are calculated via stock change factors (SCF) given by the IPCC for the topsoil (0-30 cm) and based on observational data. SCF are differentiated by different crop and management systems ($m$) reflecting different dynamics under changed in- and outflows without explicitly tracking these. Moreover SCF vary for different climatic zones ($c$) specified by the IPCC (see \@ref(append:climatemap)). The actual SOC stocks as thus calulated based on a given reference SOC stock by

\begin{equation}
SOC^{\text{target}}_{i} = \sum_{c,m} T_{c,i} \cdot SOC^{\text{ref}} \cdot F_{\text{LU}_{c,m}} \cdot F_{\text{MG}_{c,m}} \cdot F_{\text{I}_{c,m}} \cdot A_{m,i} \qquad\forall\quad t,
\label{eq:tier1}
\end{equation}

with $T_{c,i}$ being the translation matrix for grid cells $i$ into corresponding climate zones $c$. For this analysis we us the default SCF as comparison and constistency check for our more detailed budget approach.

<!-- KK: Representation of climate zones and grid cells is not consistent and has to be reformulated. -->

### 2020-06-26 SOCbudgetPaper/results.Rmd
update method part + smaller stuff in other section and result figure

```{r table, echo = FALSE}

table <- readRDS("../ResultNotebooks/Output/SFC_climatezone.rds")
knitr::kable(table$naturalstate_table, "latex", digit=2, caption="SCF compared to potential natural state")
knitr::kable(table$actualstate_table,  "latex", digit=2, caption="SCF compared to actual natural state")
knitr::kable(table$steadystate_table,  "latex", digit=2, caption="Steady state SCF compared to actual natural state")

```

>>>>

```{r table, echo = FALSE}
table <- readRDS("../ResultNotebooks/Output/SFC_climatezone.rds")
knitr::kable(table$naturalstate_table, "latex", digit=2, caption="SCF compared to potential natural state")
```

```{r table2, echo = FALSE}
table <- readRDS("../ResultNotebooks/Output/SFC_comparison.rds")
knitr::kable(table$comparison_table, "latex", digit=2, caption="SCF compared to IPCC Tier 1")
```

```{r out.width = "16cm"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/4panelfigure.png")
```
